---
title: "malaria"
format: html
---
```{r}
library(tidyverse)
```

```{r}
malaria <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2018/2018-11-13/malaria_inc.csv")
```

```{r}
#install.packages("tidymodels")
library(tidymodels)

my_data_splits <- initial_split(malaria, prop = 0.5)

exploratory_data <- training(my_data_splits)
test_data <- testing(my_data_splits)
```

```{r}
exploratory_data
  
```

How do malaria incidents change over time?
what countries have the highest rate of incidence?
What year had the highest rate of malaria?

```{r}
exploratory_data%>%
  ggplot()+
  geom_point(mapping = aes(x= Year, y = `Incidence of malaria (per 1,000 population at risk) (per 1,000 population at risk)`))
```
Need to set year as a catigorical in order to use a better graph.

```{r}
exploratory_data%>%
  group_by(Year)%>%
  summarise(mean(`Incidence of malaria (per 1,000 population at risk) (per 1,000 population at risk)`))
```
Mean incidents for each year. Decreaseing mean as time passes.

```{r}
exploratory_data%>%
  ggplot()+
  geom_boxplot(mapping = aes(x= Code, y = `Incidence of malaria (per 1,000 population at risk) (per 1,000 population at risk)`))

```
This is not useful but I am not sure how to clean it up. There are 100 contries in the total data plus some un catogrised data.


Hypothesis
Cases of malaria will decrease on average over the years.
2000 will have the most cases of malaria. 
